Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-27 response_to_referees.tex

\editor{The comments and recommendations from two expert reviewers are now available for your manuscript. These reviewers judged the reported discoveries to be of medium significance, and the potential scientific impact of your work to be medium/high. They found that the manuscript needs improvement in text and additional data analysis. Editors generally agree with their concerns and recommendations, which led to a designation of medium/high priority. }

\response{
Thank you very much for handling our manuscript and providing these two constructive reviews that have helped us to improve the paper.
}

\editor{Note that many manuscripts receiving medium priority based on reviewer comments are not accepted by the Board of Editors. If appropriate, the board may invite a resubmission following the rejection, which is intended to enable you to improve the manuscript towards receiving a high or top priority. The authors should pay close attention to the detailed review comments and address each comment with significant improvements.}

\response{We have used the revision to include additional analysis and improve the presentation. We hope that you and the reviewers now feel that our manuscript makes a major contribution towards understanding what aspects of influenza virus evolution are predictable, and which ones are not. Given the major public health importance of influenza and the fact that influenza viruses are one of the very few systems where adaptive evolution can be observed in real time, we feel that our results are of very high relevance not only to the community studying influenza virus evolution but also those studying rapid adaptive evolution more generally.
}

@@

\refa{1. The prediction problem is defined for each mutation: based on the
tracked frequency trajectory, can the future (fixation, loss, or
polymorphism) of the mutation be predicted. Such formulation has been
previously proposed by Illingworth and Mustonen, (eg. Genetics 2011,
Plos Pathogens 2012). By averaging over all amino-acid substitutions,
the authors show that the fates of mutations are not determined by the
value of the starting frequency.}

\response{We thank the reviewer for pointing out the work by Illingworth and Mustonen, which is indeed very relevant and complementary to our approach.
We discuss their relation below. }

\refa{1.1. Mutations in influenza are highly nested, with a substantial
hitch-hiking, and no attempt is made to disentangle such dependencies when counting the mutations.}

\response{
The reviewer is correct that our original manuscript did not attempt to account for nested mutations and we agree that this should have been done. \\
We added a new section in the Supplementary Material (``Correcting for nested trajectories'') where investigate the extent to which nesting affects our results.
We cluster trajectories of mutations that partly appear on the same strains to various degrees, and trajectories corresponding to mutations always or often appearing on the same strains are then counted as one ``effective" trajectory.
Our analysis shows that nesting of mutations does not change our results significantly.
We therefore left the analysis presented in the main text unchanged and refer to the supplement for a discussion of this issue.
}

@@

\refa{1.2 Despite the more general formulation in the beginning, this
approach makes use only of the last time point in the trajectory,
rather than the full trace (contrary to the work of Illingworth\&
Mustonen). I find the observed neutral-like statistics not surprising
for such a limited data input, which doesn't capture past frequency dynamics. Therefore, these conclusions should be
revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}

\response{
	Illingworth and Mustonen (I\&M) presented a careful analysis to what extent trajectories can be fit using a model assuming a single fitness coefficient, or a model that differentiates different background fitnesses.
	Our results are consistent with their finding that a single parameter model (which only allows monotonic trajectories) does not capture the majority of trajectories.
	Beyond that, our approach and aim differ from I\&M.
	We investigate \emph{a priori} features of trajectories that potentially have information on whether the mutation fixes, not whether the trajectories can be parametrized by a model.
	Fitting a model to a partial trajectory to predict fitness is not expected to work: I\&M showed that subsequent interference has a substantial effect and our results indicate that the speed at which the mutation rises does not predict fitness.

	Furthermore, I\&M estimate that fixing mutations have a much higher intrinsic fitness than dying mutations while we show that very little such information is available a priori.
	The retrospective fit to trajectories tends to assign a positive fitness effect to fixing (and hence on average rising) mutations.
	This is why we resort to a model free approach to ask what a priori information can help to determine which mutations fix or die.

    We also would like to point out that we see systematic differences between H3N2 and H1N1pdm.
    While H1N1pdm shows a fixation behavior that is compatible with interfering selective sweeps, H3N2 does not.
    H3N2 is typically regarded as the more rapidly adapting influenze lineage with the most "spindly" phylogeny.
    We therefore do think that our observations are surprising to many people.

@@

\refa{2. The authors examined the predictive power of one predictive
method, the LBI. However, they did not do a systematic comparison of
the different methods that they cite (Morris et al, 2018), which differ in prediction
targets and methods. Therefore, the general conclusions about
predictability, eg. on page 8, in line 47, and on page 9, line 48 are
too sweeping and should be made precisely for those methods looked at
in detail.}

\response{The reviewer is right to point out that our original manuscript did not perform a systematic comparison of different prediction methods.

In a recent model-driven effort (Huddleston et al), we investigate the power of sequence based, antigenically informed, experimentally informed, or phylogenetic (LBI) scores to predict future influenza virus populations.
The LBI was among the highest performing models and we hence used this score in the original manuscript.
But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material ("Ability of fitness models to predict fixation").
These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
We believe this addition has made the manuscript stronger and we thank both referees for the suggestion.

We also agree with the reviewer that these different methods differ in prediction targets: our results indicate that most of them are not predictive of frequency dynamics and fixation, but they remain predictive of the composition of the future population.
In consequence, we have toned down our conclusions about predictability. \pierre{I say that because I changed p8 l47. I have not touched p9 l48 because the new appendix shows it is a correct statement.}
}
